---
output: github_document
---

# Different ways to save your plots?

R/Rstudio and Giotto provide different ways to save your spatial data:  


- see also FAQ [How to create global instructions and show or save your created plots?](./getting_started_instructions.html)  

- All parameters available to save_param are found by running **showSaveParameters()**  

### Examples :  

#### 1. standard R way
```{r, eval=FALSE}
pl = spatPlot(mygobject, cell_color = 'cell_types')
pdf(file = 'path/to/save/to/plot.pdf')
print(pl)
dev.off()
```


#### 2. save plot directly to your designated folder according to the provided instructions  
See also the [Giotto Instruction]() information page.
```{r, eval=FALSE}
spatPlot(mygobject, 
         cell_color = 'cell_types', 
         save_plot = TRUE)
```


#### 3. save plot directly, but overwrite default parameters
See also the [Giotto Instruction]() information page.  
Run the command **showSaveParameters()** to see all available parameters.   
```{r, eval=FALSE}
spatPlot(mygobject, 
         cell_color = 'cell_types', 
         save_plot = TRUE,
         save_param = list(save_folder = 'my_subfolder', 
                           save_name = 'my_name', 
                           save_format = 'png', 
                           units = 'in'))
```


#### 4. just view the plot
See also the [Giotto Instruction]() information page.  
Set save_plot & return_plot to FALSE and show_plot to TRUE.
```{r, eval=FALSE}
spatPlot(mygobject, 
         cell_color = 'cell_types', 
         save_plot = F, return_plot = F, show_plot = T)
```


#### 5. just save the plot: FASTEST for large datasets!
See also the [Giotto Instruction]() information page.  
Set save_plot & return_plot to FALSE and show_plot to TRUE.
```{r, eval=FALSE}
# only saves the plot
spatPlot(mygobject, 
         cell_color = 'cell_types', 
         save_plot = TRUE, return_plot = FALSE, show_plot = TRUE)

# only saves the plot, but launches the saved plot in a window
spatPlot(mygobject, 
         cell_color = 'cell_types', 
         save_plot = TRUE, return_plot = FALSE, show_plot = TRUE,
         save_param = list(show_saved_plot = TRUE))
```
